\begin{center}
{\bf 作者简介}
\end{center}

Norm Matloff 博士is a professor of computer science at the University of
California at Davis, and was formerly a professor of statistics at that
university. He is a former database software developer in Silicon
Valley, and has been a statistical consultant for firms such as the
Kaiser Permanente Health Plan.

Dr. Matloff was born in Los Angeles, and grew up in East Los Angeles and
the San Gabriel Valley. He has a PhD in pure mathematics from UCLA,
specializing in probability theory and statistics.  He has published
numerous papers in computer science and statistics, with current
research interests in parallel processing, statistical computing,
and regression methodology.

Prof. Matloff is a former appointed member of IFIP Working Group 11.3,
an international committee concerned with database software security,
established under UNESCO.  He was a founding member of the UC Davis
Department of Statistics, and participated in the formation of the UCD
Computer Science Department as well.  He is a recipient of the
campuswide Distinguished Teaching Award and Distinguished Public Service
Award at UC Davis.

Matloff 博士是两本书的作者，并编写了大量广泛使用的网络教程，涉及 Linux 和 python 语言。
他和 Peter Salzman 博士是《软件调试的艺术》\footnote{The Art of Debugging with GDB, DDD, and
Eclipse，中文版已由人民邮电出版社出版}一书的作者。Matloff 博士
关于 R 语言的 《R 语言编程艺术》\footnote{The Art of R Programming，
中文版已由机械工业出版社出版}一书已于2011年出版。他的新书， {\it Parallel
Computation for Data Science}会于2014年出版。
他还写作了很多开源图书，包括{\it From Algorithms
to Z-Scores: Probabilistic and Statistical Modeling in Computer Science}
（\url{http://heather.cs.ucdavis.edu/probstatbook}），和《并行机器编程》
（\url{http://heather.cs.ucdavis.edu/~matloff/ParProcBook.pdf}）。
